\bf A hierarchical failure-time model for observational data exhibiting infant-mortality and wearout failure modes

### 2017-10-05 paper/paper.tex
End of day commit to paper

Probability plotting is a simple method to assess the adequacy of (log)location-scale families of probability models to the data.  After properly transforming the axes of a plot, graphing an empirical estimate of fraction failing as a function of time along with simultaneous confidence bands provides a visual check for distributional goodness of fit.  We applied this method using the Kaplan-Meier non-parametric estimate of the empirical cdf.  With left truncation, however, the standard Kaplan-Meier estimator is biased, so we used an adjusted version due to\citet[Chapter 11, ]{meeker} (see Appendix for a detailed description).

@@

Next, we model these parameters themselves as random variables, varying across the different sub-populations.  Let
\begin{equation*}
\sigma_{1g} \ind \op{Log-normal} \left( \eta_{\sigma_1}, \tau^2_{\sigma_1} \right)
\end{equation*}
\begin{equation*}
\sigma_{2g} \ind \op{Log-normal} \left( \eta_{\sigma_2}, \tau^2_{\sigma_2}\right)\op{T}\left(0, 1\right)
\end{equation*}
\begin{equation}
\label{eq:hier-model}
t_{p_{1}g} \equiv \mu_{1g} + \sigma_{1g}\,\Phi^{-1}(p_1)  \ind \op{Log-normal} \left(\eta_{t_{p_1}}, \tau^2_{t_{p_1}}\right)
\end{equation}
\begin{equation*}
t_{p_{2}g} \equiv \mu_{2g} + \sigma_{2g}\,\Phi^{-1}(p_2)  \ind \op{Log-normal} \left(\eta_{t_{p_2}}, \tau^2_{t_{p_2}}\right)
\end{equation*}
\begin{equation*}
\pi_g \ind \op{Logit-normal}(\eta_\pi, \tau^2_\pi).
\end{equation*}

@@

Figure \ref{fig:fig2} provides a visual comparison for Models 1 through 4 by plotting the pointwise posterior median of time to failure. The large left panel shows the Kaplan-Meier nonparametric estimate of the same. The Kaplan-Meier estimates on the right side of Figure \ref{fig:fig2} shows that the set of estimated failure curves for all four models that we fit. The smaller plots correspond to Models 1 through 4.  Note that while all the paths for the posterior estimates necessarily pass through the origin, the nonparametric curves do not, but only because, for the Kaplan-Meier estimator, inference is entirely conditional on survival up to left-most left-trunctation time; to make the plot comparable, we use an adjustment based on a parametric model (Model 4).

@@

We can also compare posterior estimates of time to failure under Models 1-4 for each drive-model individually.  For these comparisons, we calculate, pointwise, the posterior median of proportion failing over a grid. That is, for every model, and for a fixed set, $\tilde{t}=\{\tilde{t}_1,\ldots,\tilde{t}_M\}$, we compute 
\begin{equation}
\op{median}\left\{\op{GLFP}\left(\tilde{t}_m|\pi_{g}^{(s)},\mu_{1g}^{(s)},\sigma_{1g}^{(s)},\mu_{2g}^{(s)},\sigma_{2g}^{(s)}\right);s=1,\ldots,S \right\},
\end{equation}
where $\theta^{(s)}$ is the $s^{th}$ posterior sample from $p(\theta|y)$.
%Unlike maximum likelihood, by including the hyperparameters in the probability model, there is a data-dependent penalization that results in shrinkage toward a common overall model.

@@

For Drive-model 14, Models 1-3 produce lower estimates of the proportion of drives failing, and for Drive-model 40, they produce higher estimates.  On the other hand, Model 4 follows the nonparametric estimate quite closely across stages of lifetime.  The differences between Models 3 and 4 and Models 1 and 2 are visually quite apparent.  Differentiating between Model 3 and 4 is less clear; for instance, the estimates from Model 3 and 4 are nearly identical for Drive-model 40.

@@

An obstacle remains, in that the censoring mechanism itself is not fully observed. Our admittedly simple approach is to assume that failed units would have been the censored had they survived until the latest censoring or failure time for that drive model. We generated 19 replicates for each drive model. This gives a 1/20 chance under the hypothesis of no difference that the Kaplan-Meier estimate based on the observed data would be ``most extreme" relative all the replicated  from the those one and is not too many so as to produce an overly cluttered plot. Our method for generating replicate data sets, $y_{rep,g}^{(s)}$, $s = 1 \dots 19$, for each drive-model $g=1,\ldots,44$, is as follows:
\begin{enumerate}
\item Simulate $\tilde{t}_{rep,g,i}^{(s)} \sim \op{GLFP}(\pi_{g}^{(s)},\mu_1^{(s)},\sigma_1^{(s)},\mu_{2g}^{(s)}, \sigma_{2g}^{(s)})$ where each set of parameters, $\left( \pi_{g}^{(s)},\mu_1^{(s)},\sigma_1^{(s)},\mu_{2g}^{(s)}, \sigma_{2g}^{(s)} \right)$, is a random draw from the full posterior distribution.
\item Repeat step 1 until $\tilde{t}_{rep,g,i} > t_{g,i}^L$.
\item If $c_{g,i}=0$, let $c_{rep,g,i}^{(s)}=\max \{t_{g,i}: i \ge 1\}$, otherwise let $c_{rep,g,i}^{(s)}=t_{g,i}$.\\
If $\tilde{t}_{rep,g,i}^{(s)}>m_{g,i}$, set $t_{rep,g,i}^{(s)}=c_{rep,g,i}^{(s)},\; c_{rep,g,i}^{(s)}=1$, otherwise set $t_{rep,g,i}^{(s)}=\tilde{t}_{rep,g,i}^{(s)},\; c_{rep,g,i}^{(s)}=0$.
\end{enumerate}

\begin{figure}[H]
\includegraphics[width=\textwidth]{ppcheck-sample}
\caption{Nonparametric estimates of the proportion failing for a representative subset of drive-models. Both the original data (\textbf{bold} line) and 19 ``replicated" data sets from the posterior predictive distribution (\textit{dashed} lines) are shown.}
\label{fig:post-pred-KM}
\end{figure}

We can compare the 19 simulated data sets $y_{rep,g, i}$ to the observed data $y_{g,i}$ by using them to re-estimate the Kaplan-Meier nonparametric estimator. The resulting plots are displayed in Figure \ref{fig:post-pred-KM}. Only a representative sample of plots is shown here for brevity (see the Supplementary Material for the complete set of drive-models).  For most drive-models, the estimates based on the observed data look similar to those based on the replicated data.  There are a few drive-models, however where this is not the case. Drive-model 14 shows a discrepancy between the GFLP fit and the Kaplan-Meier fit in the right tail. In particular, the GLFP predictions appear conservative, overpredicting the proportion failing after about 2.5 years. Due to the large number of failures for this Drive-model, the posterior for the hierarchical model is similar to the stand-alone model. Other drive-models showing a lack of fit are 4 and 34. We first address Drive-model 4; our diagnosis of 34 is analogous, so the following comments extend to that Drive-model as well.

To diagnose the cause of the discrepancies between the observed data for Drive-model 4 and its posterior predictive distribution, we consider the event plot for a random sample of units for this Drive-model while looking at estimates of lifetime (Figure \ref{fig:ex-mod-4}). From these plots, we can see that the discrepancy between the posterior and the nonparametric estimate is driven by failure among a relatively small subset of units, namely those with left-truncation times smaller than 2500 hours. Further investigation reveals that, if we exclude the newest 75 drives, those with truncation times less than 800 hours, the nonparametric estimate falls within the posterior 90\% pointwise credible band. This situation suggests that, despite belonging to the same Drive-model, there may be something different about the newer drives leading to worse reliability.

@@

Figure 8 shows a similar pair of plots for Drive-model 43 illustrate (Figure \ref{fig:ex-mod-43}). The data for this Drive-model are heavily truncated; the observed units had run for quite some time before data were made available with the earliest left-truncation time at 12,189 hours. In Figure \ref{fig:ex-mod-43} we see that the lack of data for this Drive-model during early life results in a diffuse posterior centered around the what we call the ``global mean." The global mean is corresponds to $\op{GFLP}$ parameters taking the value of the hierarchical means, i.e., $\pi=\eta_{\pi}, \, \mu_2=\eta_{t_2} - \eta_{\sigma_2}\Phi^{-1}(.2)$ and $\sigma_2= \eta_{\sigma_2})$. This continues until about 10,000 hours at which time the posterior diverges from the global model, arriving at a compromise between it and the nonparametric estimate.

@@

\appendix
\section{Truncation adjusted nonparametric estimate of lifetime}
\label{sec:trunc-adj}
We first start with a non-parametric estimate of the empirical cdf for each drive-model using the Kaplan-Meier estimator.  With left truncation, however, the standard Kaplan-Meier estimator for drive-model $d$, denoted by
$\widehat{F_d(t)}_{KM}$, is conditional on survival up to
$t_{d,\text{min}}^L$, the shortest reported running time of all units
of drive-model $d$ for which records are available. To produce
unconditional estimates, we adapt the adjustment method outlined by \citet[Chapter 11]{meeker}.  For each drive-model we select
$t_{d,\text{min}}^L$, the smallest left truncated time in the sample.
By sampling from the full posterior distribution, since
$\Pr(T>t_{d,\text{min}}^L|\theta_d)$ (the probability a hard drive has
survived up to $t_{d,\text{min}}^L$) is a function of the model
parameters, we can easily compute its posterior median,
$\widehat{A}_{\text{med}} = \widehat{\Pr}(T>t_{d,\text{min}}^L|\theta_d)$. We compute the adjusted estimate by

>>>>

Probability plotting is a simple method to assess the adequacy of (log)location-scale families of probability models to the data.  After properly transforming the axes of a plot, graphing an empirical estimate of fraction failing as a function of time along with pointwise confidence bands provides a visual check for distributional goodness of fit.  We applied this method using the Kaplan-Meier nonparametric estimate of the empirical cdf.  With left truncation, however, the standard Kaplan-Meier estimator is biased, so we used an adjusted version due to\citet[Chapter 11, ]{meeker} (see Appendix for a detailed description).

@@

Next, we model these parameters themselves as random variables, varying across the different sub-populations.  Let
\begin{equation*}
\sigma_{1g} \ind \op{Log-normal} \left( \eta_{\sigma_1}, \tau^2_{\sigma_1} \right)
\end{equation*}
\begin{equation*}
\sigma_{2g} \ind \op{Log-normal} \left( \eta_{\sigma_2}, \tau^2_{\sigma_2}\right)\op{T}\left(0, 1\right)
\end{equation*}
\begin{equation}
\label{eq:hier-model}
t_{p_{1}g} \equiv \exp\left(\mu_{1g} + \sigma_{1g}\,\Phi^{-1}(p_1)\right)  \ind \op{Log-normal} \left(\eta_{t_{p_1}}, \tau^2_{t_{p_1}}\right)
\end{equation}
\begin{equation*}
t_{p_{2}g} \equiv \exp\left(\mu_{2g} + \sigma_{2g}\,\Phi^{-1}(p_2)\right)  \ind \op{Log-normal} \left(\eta_{t_{p_2}}, \tau^2_{t_{p_2}}\right)
\end{equation*}
\begin{equation*}
\pi_g \ind \op{Logit-normal}(\eta_\pi, \tau^2_\pi).
\end{equation*}

@@

Figure \ref{fig:fig2} provides a visual comparison for Models 1 through 4 by plotting the pointwise posterior median of time to failure. The large left panel shows the Kaplan-Meier estimate of the same. The Kaplan-Meier estimates on the right side of Figure \ref{fig:fig2} shows that the set of estimated failure curves for all four models that we fit. The smaller plots correspond to Models 1 through 4.  Note that while all the paths for the posterior estimates necessarily pass through the origin, the Kaplan-Meier estimates do not, because, for the Kaplan-Meier estimator, inference is entirely conditional on survival up to left-most left-trunctation time. To make the plot comparable, we use an adjustment based on a parametric model (Model 4).

@@

We can also compare posterior estimates of time to failure under Models 1-4 for each drive-model individually.  For these comparisons, we calculate, pointwise, the posterior median of proportion failing over a grid. That is, for every model, and for a fixed set, $\tilde{t}=\{\tilde{t}_1,\ldots,\tilde{t}_M\}$, we compute 
\begin{equation}
\label{pointwise-medians}
\op{median}\left\{\op{GLFP}\left(\tilde{t}_m|\pi_{g}^{(s)},\mu_{1g}^{(s)},\sigma_{1g}^{(s)},\mu_{2g}^{(s)},\sigma_{2g}^{(s)}\right);s=1,\ldots,S \right\},
\end{equation}
where $\theta^{(s)}$ is the $s^{th}$ posterior sample from $p(\theta|y)$.
%Unlike maximum likelihood, by including the hyperparameters in the probability model, there is a data-dependent penalization that results in shrinkage toward a common overall model.

@@

For Drive-model 14, Models 1-3 produce lower estimates of the proportion of drives failing, and for Drive-model 40, they produce higher estimates.  On the other hand, Model 4 follows the Kaplan-Meier quite closely across stages of lifetime.  The differences between Models 3 and 4 and Models 1 and 2 are visually quite apparent.  Differentiating between Model 3 and 4 is less clear; for instance, the estimates from Model 3 and 4 are nearly identical for Drive-model 40.

@@

An obstacle remains, in that the censoring mechanism itself is not fully observed. Our admittedly simple approach is to assume that failed units would have been the censored had they survived until the latest censoring or failure time for that drive model. We generated 19 replicates for each drive model. This gives a 1/20 chance under the hypothesis of no difference that the Kaplan-Meier estimate based on the observed data would be ``most extreme" relative all the replicated  from the those one and is not too many so as to produce an overly cluttered plot. Our method for generating replicate data sets, $y_{rep,g}^{(s)}$, $s = 1 \dots 19$, for each drive-model $g=1,\ldots,44$, is as follows:
\begin{enumerate}
\item Simulate $\tilde{t}_{rep,g,i}^{(s)} \sim \op{GLFP}(\pi_{g}^{(s)},\mu_1^{(s)},\sigma_1^{(s)},\mu_{2g}^{(s)}, \sigma_{2g}^{(s)})$ where each set of parameters, \\$\left( \pi_{g}^{(s)},\mu_1^{(s)},\sigma_1^{(s)},\mu_{2g}^{(s)}, \sigma_{2g}^{(s)} \right)$, is a random draw from the full posterior distribution.
\item Repeat step 1 until $\tilde{t}_{rep,g,i} > t_{g,i}^L$.
\item If $c_{g,i}=0$, let $c_{rep,g,i}^{(s)}=\max \{t_{g,i}: i \ge 1\}$, otherwise let $c_{rep,g,i}^{(s)}=t_{g,i}$.\\
If $\tilde{t}_{rep,g,i}^{(s)}>m_{g,i}$, set $t_{rep,g,i}^{(s)}=c_{rep,g,i}^{(s)},\; c_{rep,g,i}^{(s)}=1$, otherwise set $t_{rep,g,i}^{(s)}=\tilde{t}_{rep,g,i}^{(s)},\; c_{rep,g,i}^{(s)}=0$.
\end{enumerate}

\begin{figure}[H]
\includegraphics[width=\textwidth]{ppcheck-sample}
\caption{Adjusted Kaplan-Meier estimates of the proportion failing for a representative subset of drive-models. Both the original data (\textbf{bold} line) and 19 ``replicated" data sets from the posterior predictive distribution (\textit{dashed} lines) are shown.}
\label{fig:post-pred-KM}
\end{figure}

We can compare the 19 simulated data sets $y_{rep,g, i}$ to the observed data $y_{g,i}$ by using them to re-estimate the Kaplan-Meier estimator. The resulting plots are displayed in Figure \ref{fig:post-pred-KM}. Only a representative sample of plots is shown here for brevity (see the Supplementary Material for the complete set of drive-models).  For most drive-models, the estimates based on the observed data look similar to those based on the replicated data.  There are a few drive-models, however where this is not the case. Drive-model 14 shows a discrepancy between the GFLP fit and the Kaplan-Meier fit in the right tail. In particular, the GLFP predictions appear conservative, overpredicting the proportion failing after about 2.5 years. Due to the large number of failures for this Drive-model, the posterior for the hierarchical model is similar to the stand-alone model. Other drive-models showing a lack of fit are 4 and 34. We first address Drive-model 4; our diagnosis of 34 is analogous, so the following comments extend to that Drive-model as well.

To diagnose the cause of the discrepancies between the observed data for Drive-model 4 and its posterior predictive distribution, we consider the event plot for a random sample of units for this Drive-model while looking at estimates of lifetime (Figure \ref{fig:ex-mod-4}). From these plots, we can see that the discrepancy between the posterior and Kaplan-Meier estimates is driven by failures among a relatively small subset of units, namely those with left-truncation times smaller than 2500 hours. Further investigation reveals that, if we exclude the newest 75 drives, those with truncation times less than 800 hours, the Kaplan-Meier estimate falls within the posterior 90\% pointwise credible band. This situation suggests that, despite belonging to the same Drive-model, there may be something different about the newer drives leading to worse reliability.

@@

Figure 8 shows a similar pair of plots for Drive-model 43 illustrate (Figure \ref{fig:ex-mod-43}). The data for this Drive-model are heavily truncated; the observed units had run for quite some time before data were made available with the earliest left-truncation time at 12,189 hours. In the bottom half of Figure \ref{fig:ex-mod-43} we contrast the posterior median for Drive-model 43 to what we call the `global average.' We get this by setting the $\op{GLFP}$ parameters equal to the median of their hierarchical distribution, i.e., $\pi=\op{logit}^{-1}(\eta_{\pi}),\, \mu_1=\mu_1,\, \sigma_1=\sigma_1,\, \mu_2=\eta_{t_2} - m_{\sigma_2}\Phi^{-1}(.2)$ and $\sigma_2= m_{\sigma_2}$, where $m_{\sigma_2}=\Phi_{Log-Normal}^{-1}[0.5 \cdot \Phi_{Log-Normal}(1|\eta_{\sigma_2},\tau_{\sigma_2})|\eta_{\sigma_2}, \tau_{\sigma_2}]$ is the population median of the $\sigma_{2g}$. The curve was estimated using pointwise medians as in Equation \ref{pointwise-medians}. We can see that the lack of data for this Drive-model during early life results in a diffuse posterior centered around the global average until about 10,000 hours when they diverge. Although the posterior is close to the adjusted Kaplan-Meier, it shows shrinkage toward the global average.

@@

\appendix
\section{Truncation adjusted Kaplan-Meier estimate of lifetime}
\label{sec:trunc-adj}
We first start with a non-parametric estimate of the empirical cdf for each drive-model using the Kaplan-Meier estimator.  With left truncation, however, the standard Kaplan-Meier estimator for drive-model $d$, denoted by
$\widehat{F_d(t)}_{KM}$, is conditional on survival up to
$t_{d,\text{min}}^L$, the shortest reported running time of all units
of drive-model $d$ for which records are available. To produce
unconditional estimates, we adapt the adjustment method outlined by \citet[Chapter 11]{meeker}.  For each drive-model we select
$t_{d,\text{min}}^L$, the smallest left truncated time in the sample.
By sampling from the full posterior distribution, since
$\Pr(T>t_{d,\text{min}}^L|\theta_d)$ (the probability a hard drive has
survived up to $t_{d,\text{min}}^L$) is a function of the model
parameters, we can easily compute its posterior median,
$\widehat{A}_{\text{med}} = \widehat{\Pr}(T>t_{d,\text{min}}^L|\theta_d)$. We compute the adjusted estimate by
